White Dwarf Mergers on Adaptive Meshes\\ III. Mergers

### 2016-09-12 papers/wdmerger_III/paper.tex
Update the description of the hybrid advection scheme

%==========================================================================
% Numerical Methods
%==========================================================================
\section{Numerical Methods}

\subsection{Hybrid Advection Scheme}

It is well known that there is a trade-off between conservation of linear
momentum and conservation of angular momentum in grid-based, Eulerian codes
such as \castro. Although both are true statements analytically, we cannot
numerically guarantee that both hold. Consequently the choice of which
physical quantity to conserve must be motivated by analysis of the particular
problem to be done. Ideally a simulation code will be capable of evolving the
equations for both the linear momentum and the angular momentum, and be
able to switch between them as appropriate. Prior to this work, \castro\
already had the standard method implemented for evolving the linear momentum.
To solve the need for an angular momentum evolution, we adopt the ``hybrid''
advection scheme presented by \cite{byerly:2014} (see also \cite{motl:2002} for
a similar equation set). Their approach is appropriate for physical situations
where there is a dominant angular momentum axis and it is desirable to conserve
that component as accurately as possible. This is certainly applicable for
studies of the stability of binary stellar systems. This section contains a
discussion of our implementation of that scheme in \castro. For the sake
of simplicity, let us assume that the rotation axis of the binary system is the $z$
axis and so the WDs orbit in the $xy$-plane. The core of the method is that instead
of solving the Euler equations for the linear momenta $(\rho u)$ and $(\rho v)$,
we instead solve the corresponding equations for the radial momentum
$s_R \equiv \rho v_R$ (where $v_R$ is the radial velocity with respect to the
rotation axis) and the angular momentum with respect to the $z$-axis,
$\ell_z \equiv R\rho v_\phi$ (where $v_\phi$ is the azimuthal velocity).
The equation for the linear $z$-momentum is unchanged. We want to be able
to solve these equations while still operating in our preferred Cartesian
coordinate system. \cite{byerly:2014} present these equations as the following;
for the moment we neglect inclusion of external source terms such as gravity and rotation:
\begin{align}
  \frac{\partial(s_R)}{\partial t} + \nabla \cdot (s_R \mathbf{u}) &=
    - \frac{1}{R}\left(x \frac{\partial}{\partial x} + y \frac{\partial}{\partial y}\right) p
    + \frac{\ell_z^2}{\rho R^3} \label{eq:radial-momentum}\\
    \frac{\partial(\ell_z)}{\partial t} + \nabla \cdot (\ell_z \mathbf{u}) &=
  \left(y\frac{\partial}{\partial x} - x \frac{\partial}{\partial y}\right) p \label{eq:angular-momentum}
\end{align}  
Here $\mathbf{u}$ is the ordinary linear momentum used to advect any fluid quantity
on the grid, and $R = \sqrt{x^2 + y^2}$, where the coordinates $x$ and $y$ are defined
relative to the axis of rotation, and the origin is located on a zone corner
so that singularities are avoided. Note that $v_R = (1 / R)(x u + y v)$, and
$v_\phi = (1/R)(x v - y u)$. These equations can be straightforwardly derived from
the Euler equations for linear momentum by appropriate multiplication of $x$, $y$,
$x / R$, and $y/R$, and subsequent algebraic manipulation. We now rewrite this in the
way that we actually discretize them for use in the code (by use of the product rule
to rearrange the terms with derivatives in them):
\begin{align}
  \frac{\partial(s_R)}{\partial t} &= -\nabla \cdot (s_R \mathbf{u})
    - \left(\frac{\partial(p (x/R))}{\partial x} + \frac{\partial(p (y/R))}{\partial y}\right) \notag \\
    &\hspace{0.175in} + \frac{1}{\rho R^3}\left(\ell_z^2 + p\rho R^2\right) \label{eq:radial-momentum-rearranged}\\
    \frac{\partial(\ell_z)}{\partial t} &= -\nabla \cdot (\ell_z \mathbf{u}) - \left(\frac{\partial (-py)}{\partial x} + \frac{\partial(px)}{\partial y}\right).\label{eq:angular-momentum-rearranged}
\end{align}
It is clear that the second term on the right-hand-side in each case can be swept up
into the flux terms in the same way that the pressure term for the linear momenta
is often added to the fluxes, to make the conservation form explicit. The remainder
is the last term on the right-hand-side of the radial momentum equation, which
exists as a cell-centered source term (which we deal with using a predictor-corrector
method similar to how we implement external forcing like gravity and rotation; see
\cite{castro} for details). It is important to observe that in this formulation
the radial momentum equation is \textit{not} a conserved quantity: it should and
will in general change with time over the course of the simulation. The above form,
absent of source terms, makes this clear. The angular momentum term in 
\autoref{eq:angular-momentum} can be thought of physically as expressing the fact
that a particle with some initial angular momentum will have its radial momentum
increase with time as it moves away from the origin, and can be thought of geometrically
as expressing the curved nature of the underlying coordinates for the radial momentum
\citep{motl:2002}. Note that while the version in \autoref{eq:radial-momentum-rearranged}
is more convenient for our hydrodynamics methodology, the pressure-gradient terms
in the first version are the ones appropriate for a radial field and so no physical meaning
should be ascribed to the pressure term at the end of \autoref{eq:radial-momentum-rearranged}.

@@

To implement this method in \castro, we desired an approach that was flexible
and would leave as little imprint on the code structure as possible.
To this end, what we have done is to add three new ``hybrid'' momentum state
variables, corresponding to the radial momentum, angular momentum, and the linear
momentum component that is perpendicular to both. Whenever we update the normal
momentum state variables with hydrodynamics fluxes, we add the analogous fluxes
to the hybrid momenta -- that is, using the edge state values determined by the
final multi-dimensional Riemann solve, we construct the value of the hybrid
momenta on that zone edge, and then allow it to be transported as usual by the
advective velocity $\mathbf{u}$. When we update the normal momenta with an
external forcing $\mathbf{F}$ such as gravity or rotation, with $x$ and $y$
components $F_x$ and $F_y$, we apply the update to the hybrid momenta as well:
\begin{align}
  \left.\frac{\partial(s_R)}{\partial t}\right|_{F} &= -F_x \frac{x}{R} - F_y\frac{y}{R} \label{eq:force-radial}\\
  \left.\frac{\partial(\ell_z)}{\partial t}\right|_{F} &= F_x\, y - F_y\, x.\label{eq:force-angular}
  \end{align}
Then, when we are in a phase of the evolution when we want to conserve angular
momentum, we make one additional change: at the end of a (single-level) advance,
we recompute the normal momenta so that they are fully consistent with the hybrid
momenta. This is all that is necessary to have a calculation that conserves angular
momentum over the course of the simulation, without changing any of the
core infrastructure for the hydrodynamics update. For example, when we compute
the primitive variables from the conservative variables, we will still be
getting them from the normal momenta, but these normal momenta are always consistent
with the conserved angular momentum, so we get the same result as if we had done
a much more involved update where we explicitly computed the primitive variables
using the hybrid momenta. There is of course no unique choice for how to do this.
For example, we could take the approach of explicitly re-calculating the normal
momenta after any change has been made to the state. This would add significant code
complexity without a clear benefit. The most significant difference would be
that when evaluating the new-time value of source terms like rotation, the velocity
field used is slightly different than it would otherwise be; but this is a high-order
effect that does not change the core angular momentum conservation property. An
analysis of the different mechanisms for implementing this update may be an interesting
avenue for future research --- including various aspects like how to implement the
source terms in the radial momentum equation. Of course, the answer will likely
be problem-dependent. Indeed, we already faces this situation in general with the
gravitational and rotational source terms, even independent of the hybrid momentum
discussion --- the velocity field seen by the new-time rotation source terms depends
on whether we apply the gravitational force before or after the rotation force.

%==========================================================================
% Conclusions
%==========================================================================
\section{Conclusions and Discussion}\label{Sec:Conclusions and Discussion}

@@

>>>>

%==========================================================================
% Numerical Methods
%==========================================================================
\section{Numerical Methods}
\label{sec:methodology}

The numerical methodology for our problem was established in \cite{wdmergerI}.
In this section we report only on the differences between our current methodology
and the methodology in that work.

\subsection{Hybrid Advection Scheme}
\label{sec:hybrid}

It is well known that there is a trade-off between conservation of linear
momentum and conservation of angular momentum in grid-based, Eulerian codes
such as \castro. Although both quantities should be conserved analytically, we cannot
numerically guarantee that both hold. Consequently the choice of which
physical quantity to conserve must be motivated by analysis of the particular
problem to be done. In \cite{wdmergerI}, we demonstrated that numerical conservation
of linear momentum leads to white dwarf mergers that are spuriously caused by violation
of numerical conservation of angular momentum. Ideally our simulation code should be capable of evolving the
equations for both the linear momentum and the angular momentum, and be
able to switch between them as appropriate. Prior to this work, \castro\
already had the standard method implemented for evolving the linear momentum.
To solve the need for an angular momentum evolution, which could end up being more
appropriate for binary star systems than linear momentum evolution (a proposition we
will test in \autoref{sec:results}), we adopt the ``hybrid''
advection scheme presented by \cite{byerly:2014} (see also \cite{motl:2002} for
a similar equation set). Their approach is appropriate for physical situations
where there is a dominant angular momentum axis and it is desirable to conserve
that component as accurately as possible. This is certainly applicable for
studies of the stability of binary stellar systems. This section contains a
discussion of our implementation of that scheme in \castro.

For the sake of simplicity, let us assume that the rotation axis of the binary system is the $z$
axis and so the WDs orbit in the $xy$-plane. The core of the method is that instead
of solving the Euler equations for the linear momenta $(\rho u)$ and $(\rho v)$,
we instead solve the corresponding equations for the radial momentum
$s_R \equiv \rho v_R$ (where $v_R$ is the radial velocity with respect to the
rotation axis) and the angular momentum with respect to the $z$-axis,
$\ell_z \equiv R\rho v_\phi$ (where $v_\phi$ is the azimuthal velocity
and $R$ is the distance to the rotation axis).
The equation for the linear $z$-momentum is unchanged. We want to be able
to solve these equations while still operating in our preferred Cartesian
coordinate system. \cite{byerly:2014} present these equations as the following
(for the moment we neglect inclusion of external source terms such as gravity and rotation,
and we note that we have reorganized terms slightly relative to their presentation):
\begin{align}
  \frac{\partial(s_R)}{\partial t} &= -\nabla \cdot (s_R \mathbf{u})
    - \frac{1}{R}\left(x \frac{\partial p}{\partial x} + y \frac{\partial p}{\partial y}\right)
    + \frac{\ell_z^2}{\rho R^3} \label{eq:radial-momentum}\\
    \frac{\partial(\ell_z)}{\partial t} &= -\nabla \cdot (\ell_z \mathbf{u}) - \left(\frac{\partial (-py)}{\partial x} + \frac{\partial(px)}{\partial y}\right).\label{eq:angular-momentum}
\end{align}
Here $\mathbf{u}$ is the ordinary linear momentum used to advect any fluid quantity
on the grid, and $R = \sqrt{x^2 + y^2}$, where the coordinates $x$ and $y$ are defined
relative to the axis of rotation, and the origin is located on a zone corner
so that singularities are avoided. Note that $v_R = (1 / R)(x u + y v)$, and
$v_\phi = (1/R)(x v - y u)$. These equations can be straightforwardly derived from
the Euler equations for linear momentum by appropriate multiplication of $x$, $y$,
$x / R$, and $y/R$, and subsequent algebraic manipulation. The first term on the
right-hand-side is the standard advective flux term for any variable that advects with
the flow, and the state quantity inside this advective term can be derived on cell edges
using the primitive hydrodynamic variables after a Riemann solve has been performed.
The second term on the right-hand-side of \autoref{eq:angular-momentum} can be swept up
into the flux terms in the same way that the pressure term for the linear momenta
is often added to the fluxes, to make the conservation form explicit. The remainder
are the last two terms on the right-hand-side of \autoref{eq:radial-momentum}.
While maintaining second-order accuracy in time, the last term can be treated as a
cell-centered source term that we deal with using a predictor-corrector
method (similar to how we implement external forcing like gravity and rotation; see
\cite{castro} for details). The pressure gradients in the second-to-last term can be
constructed using the Riemann-solved edge states, and they are multiplied by zone
coordinates that are cell-centered. In principle this is second-order accurate in time
because the edge states are too; however, as a drawback, the pressure is unaware of the
source terms that are applied in the corrector step after a hydrodynamics update. The
benefit of this approach is that we do not need to construct a general cell-centered
gradient of the pressure using cell-centered state data, nor do we need to perform the
equation of state call that entails.

It is important to observe that in this formulation the radial momentum equation
is \textit{not} a conserved quantity: it should and will in general change with
time over the course of the simulation if the system is not in perfect rotational
equilibrium. The above form, absent of source terms, makes this clear. The angular
momentum term in \autoref{eq:radial-momentum} is effectively an outward centrifugal
force that tends to increase the radial momentum with time. This is a reflection of
the fact that mass on the domain with non-zero momentum tends to increase its radial
momentum with time. For example, consider a star initially on the $x$ axis moving
in the positive $y$ direction at constant speed: initially its radial momentum is
zero, but as its $y$ coordinate increases, more of its velocity becomes radial rather
than angular).  This term can thus also be thought of geometrically as expressing the
curved nature of the underlying coordinates for the radial momentum
\citep{motl:2002}. In the presence of a source term like gravity that can keep
such a star on an orbit at its original radius, the gravitational force acts as a
centripetal force that balances the centrifugal force, allowing the radial momentum
to remain constant with time, as it should when the orbital radius is constant.

@@

To implement this method in \castro, we desired an approach that was flexible
and would leave as little imprint on the code structure as possible.
To this end, what we have done is to add three new ``hybrid'' momentum state
variables, corresponding to the radial momentum, angular momentum, and the linear
momentum component that is perpendicular to both. Whenever we update the normal
momentum state variables with hydrodynamics fluxes, we add the analogous fluxes
to the hybrid momenta -- that is, using the edge state values determined by the
final multi-dimensional Riemann solve, we construct the value of the hybrid
momenta on that zone edge, and then allow it to be transported as usual by the
advective velocity $\mathbf{u}$. When we update the normal momenta with an
external forcing $\mathbf{F}$ such as gravity, with $x$ and $y$
components $F_x$ and $F_y$, we apply the update to the hybrid momenta as well:
\begin{align}
  \left.\frac{\partial(s_R)}{\partial t}\right|_{F} &= -F_x \frac{x}{R} - F_y\frac{y}{R} \label{eq:force-radial}\\
  \left.\frac{\partial(\ell_z)}{\partial t}\right|_{F} &= F_x\, y - F_y\, x.\label{eq:force-angular}
  \end{align}

If in addition to evolving these variables we want to use them in place of the
standard linear momenta for determining the hydrodynamic flow, only one change
is needed for the standard timestep advance scheme: after source terms are
applied to the state, we recompute the normal momenta so that they are fully
consistent with the hybrid momenta. In practice for a \castro\ advance this
synchronization of the linear and hybrid momenta occurs twice: once after the
old-time source terms and the hydrodynamic update has been applied (these all
use time-level $n$ data to construct the update, which is expected to already have
synchronized linear and hybrid momenta), and once after new-time source
terms have been applied. The use of adaptive mesh refinement implies sychronization
in two other places: after a reflux that syncs up the fluxes between coarse and
fine grids; and, after interpolation has been performed to obtain fine grid data
from a coarser grid (since the spatially interpolated data may not respect the
consistency relationship between the two momenta). In this way, the linear momenta
are essentially placeholder variables that always reflect what the current radial
and angular momenta are, but it is quite useful to express the scheme in this way
because most of the \castro\ infrastructure is already set up to use the linear
momentum state variables for calculating source terms, so the hybrid scheme requires
minimal changes to the code architecture. For example, when we compute
the primitive variables from the conservative variables in preparation for an advective
hydrodynamic update, we will still be getting them from the linear momenta,
but these linaer momenta are consistent with the conserved angular momentum,
so we get the same result as if we had modified the code to explicitly compute
the primitive variables as a function of the hybrid momenta. Consequently the
scheme presented here adds the ability to explicitly conserve angular momentum
in a simulation without a significant footprint on the simulation software.

%==========================================================================
% Results
%==========================================================================
\section{Simulation Results}\label{sec:results}

%==========================================================================
% Conclusions
%==========================================================================
\section{Conclusions and Discussion}\label{sec:Conclusions and Discussion}

@@

The authors thank Dominic Marcello and Zachary Byerly for helpful comments
regarding the hybrid advection scheme.
